BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2025-04-28 manuscript/BarnebyLives An R package to create and qc herbarium labels.rmd
workflow image slightly more legible

The formatting module has three functions.Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

@@

![The spatial extent (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map-trim.png){width=50%}

@@

# the mismatch between POW genus and not
genus_mismatch <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus) %>% 
  select(Collection_number, Scientific_name, Genus, POW_Genus) %>% 
  mutate(Situation = 
    case_when(
      Genus %in% c('Glandularia', 'Oenothera', 'Chamerion') ~ 'Preferred',
      # check Pleuraphis, i am confused on this still....
      Genus %in% c('Cryptantha', 'Ivesia', 'Lotus', 'Polygala', 'Minuartia', 'Pleuraphis',
                    'Pseudostellaria', 'Achnatherum', 'Scirpus', 'Peritoma', 'Arenaria', 
                    'Spartina', 'Polygonum', 'Atriplex', 'Cymopterus', 'Trisetum', 'Bahia', 
                   'Oryzopsis', 'Potentilla', 'Pascopyrum'
                    ) ~ 'Outdated',
      Genus %in% c('Allium', 'Vaccinium', 'Eleocharis') ~ 'Internal_Error',
      .default = 'Typo'
    ),
    Situation = if_else(Scientific_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to do so in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.

@@

LaTeX offers well-documented and detailed functionality to customize labels for individuals and institutions. 
Anecdotally, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project.

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal. 
Thematically, it's area of interest is Western North America.

>>>>

The formatting module has three functions. Two are optional; however, they are run locally and so quickly that there is no reason to skip them. 
*date_parser* parses an input date into various formats for notating collection and determination dates on labels.
*associate_dropper* silently removes the collected species from the list of associated species; however, it searches for the species to be removed using the scientific name entered initially by the user rather than returned via spell checks. 
*field_lengths* will emit messages for any fields that we suspect will create an 'overflow' on the physical label and should be truncated for clarity.

@@

![The spatial extent-or domain- (orange), and herbarium collection sites (burgundy) tested in this manuscript.](../graphics/plots/collections_map.png){width=50%}

@@

# the mismatch between POW genus and not
genus_mismatch <- data_p %>% 
  sf::st_drop_geometry() %>% 
  filter(POW_Genus != 'NOT FOUND') %>% 
  filter(POW_Genus != Genus) %>% 
  select(Collection_number, Scientific_name, Genus, POW_Genus) %>% 
  mutate(Situation = 
    case_when(
      # check Pleuraphis, i am confused on this still....
      Collection_number %in% c(
        '1231', '1243', '1245', '1248', '1251', '1263', '1299', '1363', '1373',
        '1392', '1400', '1438', '1458', '1570', '1599', '1603', '1628', '2060',
        '2094', '2185', '2233', '2249', '2249', '2258', '2264', '2291', '2421', 
        '2424', '2426', '2473', '2538', '2545', '2569', '2581', '2598', '2620', 
        '2638', '2684', '2686', '2697', '2715', '2806', '2810', '2812'
                    ) ~ 'Outdated',
      Collection_number %in% c('1236', '207', '2115', '2212', '2803') ~ 'Preferred', 
      Collection_number %in% c('2435', '2517' ) ~ 'Internal_Error',
      .default = 'Typo'
    ),
    Situation = if_else(Scientific_name == 'Oenothera caespitosa Nutt.', 'Internal_Error', Situation))

@@

While numerous tools have been developed for cleaning existing herbarium and museum records, few tools help to ensure that the data entered are accurate [@patten2024geographic].
We argue that the original collectors are the most qualified individuals to perform quality control checks and that BarnebyLives allows them to assume that responsibility in a relatively fast and streamlined format.
By utilizing both R and LaTeX and having publicly available source code on Github, this program allows users immediate familiarity with the system for troubleshooting issues and implementing upgrades and modifications in project branches.

LaTeX, a software system used for typesetting, allows users to focus on the content rather than the style of the documents rendered from it. 
However, using its default settings, it can produce aesthetically pleasing results (Figure 4). 
Additionally LaTeX offers users a wide variety of ways which they can modify labels which are under-explored in the package. 
Very good documentation of LaTeX capabilities is offered in multiple areas; for instance, via the [Overleaf](https://www.overleaf.com/learn) project. 
While the templates in the package are quite simple, LaTeX also offers the ability to use custom fonts, to alter font weights and colors, alter line spacing, to include images (e.g. dot maps) and customize labels beyond what the default templates support.

Thematically, BarnebyLives is set up to cover Western North America. 
However, the package supports the use of a 'domain' being drawn over any of the conterminous United States. 
Several of the attributes which it collects and displays on labels, relate to topics which more senior curators are interested in, i.e. the administrative information on Township Section and Range (or 'TRS'), but are considered less value in other geographic regions.  
Further several of the abiotic variables which it acquires information on: slope, aspect, and geology have long been considered prominent drivers of plant distributions in semi-arid and montane systems and warranted on a label in these types of systems, whereas curators in other regions may find this information superfluous. 
Finally, it is plausible people in other geographic areas are less interested in displaying which land management agency has jurisdiction over a collection; however in the west we believe this is useful information which may help a collector interested in revisiting a site to determine if they will require permits for access or to make new collections.

@@

# CONCLUSIONS
BarnebyLives is an R package that can be used to rapidly acquire relevant geographic and taxonomic data. 
It can also perform specialized spell checks and assorted curatorial tasks to produce both digital and analog data. 
The package relies on no licensed software, such as the Microsoft Office suite, and is suitable for install on all major operating systems  (Windows, Mac, Linux), however currently label generation support is only offered on Linux and Mac, with a small amount of use of the command line, which may be called from the Rstudio rather than a 'traditional' terminal.
